Dependent Object Types

### 2012-07-01 doc/semester/report.tex
ex section

% TODO

>>>>

We briefly summarize the salient features of the original DOT
calculus. The reader is encouraged to consult the appendices for the
full description. We will often refer to the rules of the calculus
in the remainder of this report.

The terms in DOT consists of variables $x$, $y$, $z$, lambda
abstractions $\abs x T t$, function application $\app t t'$, field
selections $t.l$ and object creation expressions $\new x c t$, where
$c$ is a constructor $T_c \ldefs{\seq{l = v}}$.

The types in DOT consists of type selections $p.L$, refinement types
$T \refine z {\seq{D}}$, function types $T \tfun T'$, type intersections
$T \tand T'$, type unions $T \tor T'$, a top type $\Top$, a bottom
type $\Bot$.

The DOT calculus is expressed using a small-step operational
semantics. The typing judgement directly or indirectly uses the
following other judgements: membership, expansion, well-formedness and
subtyping. Expansion ``flattens'' all the declarations of a type, and
membership selects a particular flattened declaration by
label. Expansion is precise. The \textsc{new} typing judgement relies
on the preciseness of expansion to check that all bounds of types
declared within the type of the constructor are consistent (i.e. that
the lower bound is a subtype of the upper bound), and that all values
declared within the type of the constructor are initialized.

\subsection{Program Examples}

\subsubsection{Basics: Booleans, Error, \ldots}
This program defines a $\mi{root}$ object with basic types
($\mi{Unit}$, $\mi{Boolean}$, $\mi{Nat}$) and values ($\mi{true}$,
$\mi{false}$, $\mi{error}$, $\mi{zero}$). For brevity, I've omitted
the code related to natural numbers. During the object creation, the
value labels such as $\mi{false}$ are all initialized.

\begin{align*}
&\mlnew {\mi{root}} {\Top \mlrefine r {\\\displaybreak[3]
&\ \Ldecl {\mi{Unit}} \Bot \Top\\
&\ \ldecl {\mi{unit}} {\Top \tfun {r.\mi{Unit}}}\\
&\ \Ldecl {\mi{Boolean}} \Bot {\Top \mlrefine z {\\
&\ \gap \ldecl {\mi{ifNat}} {(r.\mi{Unit} \tfun r.\mi{Nat}) \tfun (r.\mi{Unit} \tfun r.\mi{Nat}) \tfun r.\mi{Nat}}\\
&\gap}}\\
&\ \ldecl {\mi{false}} {{r.\mi{Unit}} \tfun {r.\mi{Boolean}}}\\
&\ \ldecl {\mi{true}} {{r.\mi{Unit}} \tfun {r.\mi{Boolean}}}\\
&\ \ldots\\\displaybreak[3]
&\ \ldecl {\mi{error}} {{r.\mi{Unit}} \tfun \Bot}\\
&}\mlldefs{\\\displaybreak[3]
&\ \mi{unit}  = \abs x \Top {\new u {\mi{root}.\mi{Unit}} u}\\
%
&\ \mi{false} = \abs u {\mi{root}.\mi{Unit}} {\\
&\ \gap \mlnew {\mi{ff}} {\mi{root}.\mi{Boolean} \mlldefs{\\\displaybreak[3]
&\ \gap\gap {\mi{ifNat}} = \abs t {\mi{root}.\mi{Unit} \tfun \mi{root}.\mi{Nat}} {\\&\gap\gap\gap\abs e {\mi{root}.\mi{Unit} \tfun \mi{root}.\mi{Nat}}\\&\gap\gap\gap\app e {\mi{root}.\mi{unit}}}\\
&\ \gap}}{
\ \gap \mi{ff}}
&}\\
%
&\ \mi{true} = \ldots\\
%
&\ \mi{error} = \abs u {\mi{root}.\mi{Unit}} {\app {\mi{root}.\mi{error}} u}\\
&\ \ldots\\
&}}{
\ldots
}
\end{align*}

\subsubsection{Polymorphic Lists}
Polymorphic lists can be expressed using an abstract type member for
the element type ($\mi{Elem}$). We can instantiate a refinement of the
list package to manipulate lists with a particular element type. We
can choose to make the lists invariant or covariant in the element
type by fixing the lower bound of the declaration to the upper bound
or to $\Bot$, respectively.

\begin{align*}
&\mlnew {\mi{genLists}} {\Top \mlrefine g {\Ldecl {\mi{ListPackage}} \Bot {\Top \mlrefine p {\\\displaybreak[3]
&\ \Ldecl {\mi{Elem}} \Bot \Top\\
&\ \Ldecl {\mi{ListOfElement}} {g.\mi{List} \refine z {\Ldecl {\mi{Elem}} {p.\mi{Elem}} {p.\mi{Elem}}}} {\\&\gap\gap\gap\gap g.\mi{List} \refine z {\Ldecl {\mi{Elem}} {p.\mi{Elem}} {p.\mi{Elem}}}}\\
&\ \Ldecl {\mi{List}} \Bot {\Top \mlrefine z {\\\displaybreak[3]
&\ \gap \ldecl {\mi{isEmpty}} {\mi{root}.\mi{Unit} \tfun \mi{root}.\mi{Boolean}}\\
&\ \gap \ldecl {\mi{head}} {\mi{root}.\mi{Unit} \tfun g.\mi{Elem}}\\
&\ \gap \ldecl {\mi{tail}} {\mi{root}.\mi{Unit} \tfun g.\mi{ListOfElem}}\\
&\gap}}\\
&\ \ldecl {\mi{nil}} {\mi{root}.\mi{Unit} \tfun g.\mi{ListOfElem}}\\
&\ \ldecl {\mi{cons}} {g.\mi{Elem} \tfun g.\mi{ListOfElem}}\\
&}}}}{
\mlnew {\mi{natLists}} {{\mi{genLists}.\mi{ListPackage}} \mlrefine p { \Ldecl {\mi{Elem}} {\mi{Nat}} {\mi{Nat}} }\mlldefs{\\\displaybreak[3]
&\ {\mi{nil}} = \ldots\\
&\ {\mi{cons}} = \ldots\\
&}}}{
\ldots
}
\end{align*}
